---
title: "Compotools Usage"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{compotools Usage}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```
### Compositional Variable Selection and Regression
\code{cdmm} is used to perform variable selection and fit a linear log-contrast model where the predictor variables are compositional data.

First, we load example data. The data includes 98 samples, each for a healthy volunteer. The column \code{bmi} is the body mass index of the person. Other columns are DNA sequencing data of the volunteers' stool samples. We are interested in identifying a subset of important genera which is associated with BMI.
```{r data, include=TRUE}
library(compotools)
data("bmi_gut") # load example data
y = bmi_gut$bmi
x = bmi_gut[,setdiff(colnames(bmi_gut), "bmi")]
```

```{r data_stat, include=TRUE}
nreads = rowSums(x) # how many reads a sample has
mean(nreads)
sd(nreads)
```
Since the number of sequencing reads varied greatly across samples, these count data should not be used directly in a standard regression analysis. So we transform them into compositonal data and then log them. As we are to use log-contrast model to analyse the data, we replace zero counts by the maximum rounding error 0·5 before transforming them.
```{r data_transform, include=TRUE}
x[x==0] = 0.5
x = x/rowSums(x)
z = log(as.matrix(x))
```
Use \code{cv.cdmm} to choose optimized tunning parameter and fit a model.
```{r cdmm, include=TRUE}
set.seed(1234)
n = nrow(z)
train_idx = sample(1:n, size = floor(n*0.7))
zt = z[train_idx,]
yt = y[train_idx]
zv = z[-train_idx,]
yv = y[-train_idx]

model = cv.cdmm(yt, zt)
model
```
Genera corresponding to none-zero coefficients are selected.
```{r corresponding_genera}
colnames(x)[model$bet != 0]
```
We use \code{cdmm} with \code{lam = 0} to train a refitted model on the selected genera.
```{r refitted_model, include=TRUE}
selected = model$bet != 0
refitted_model = cdmm(yt, zt[,selected], lam = 0)
refitted_model
```
Use refitted model to predict bmi.
```{r predict, include=TRUE, fig.width=6, fig.height=6, dpi=300}
yp <- refitted_model$int + zv[, selected] %*% refitted_model$bet
ran = range(c(yv, yp))

plot(yv, yp, 
     xlim = ran,
     ylim = ran,
     asp = 1,  
     xlab = "observed BMI", ylab = "predicted BMI",
     ) 
abline(a = 0, b = 1, lty = 2, col = "grey") 
```
### Two Samples Test for Compositional Data 

\code{cd.test} is used to assess whether there are significant differences between two compositional data samples. In this example, we test if lean people (BMI < 25) and obese ones (BMI >= 25) have the same gut microbiome composition .
```{r test}
xle = x[y<25,]
xob = x[y>=25,]

p.value = cd.test(xle, xob, paired = FALSE)
print(p.value)
```
The test gives a p value of about 0.001, indicating a marked difference between two groups.

### Covariance Estimate for Compositional Data

```{r cv.coat}
foldid = sample(1:nrow(xle)) %% 5
rcv = cv.coat(xle, foldid = foldid)
```
